---
title: "Project 1 Draft Report"
format: html
editor: visual
author: "Siddharth Venkiteswaran"
---

## Plastic Pollution

The plastic pollution data set contains information from the organization, Break Free From Plastic, and their audits on plastic pollution and the efforts to clean it up within different countries. It contains data from audits in 2019 and 2020, in addition to the parent company from where the pollution originated from. The plastic is specifically classified by recycling codes.

Looking at the cleaning that has been done to the data so far, it is clear that from the cleaning script, the columns have been set to have values of either values of characters or doubles. This applies for both 2019 and 2020. Furthermore, for 2019, there has been an str_replace operation conducted to better set the names from the file which it was collected from. Additionally, there has been a mutate operation conducted to create new variables of pp and ps, with if else statements to handle NA values within the variable pp_2.

### Initial Research Questions

1.  Which companies had the highest grand total number of plastics thrown out in 2019 and 2020?

2.  Which countries had clean up efforts that found the largest grand total of plastics in 2019 and 2020?

3.  Does the amount of plastic waste within countries that are part of the Paris Climate Agreement significantly differ from countries who are not a part of the agreement?

4.  Do countries with a higher GDP tend to contribute to rising levels of plastic waste?

```{r}
library(tidyverse)
plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')
```

### Visualization Sketches

<https://excalidraw.com/#json=buS3O8kzDji6O6ABK-REi,E2ouf_AO14QoCr69V5YTWw>

## Data Summaries Using Existing Columns

### Data Summary 1

**Question**: Which parent companies contribute the most total plastic waste across all countries, and what share of the total recorded waste does each company account for?

```{r}
overall_total <- sum(plastics$grand_total, na.rm = TRUE)

company_share <- function(x, overall_total) {
  sum(x, na.rm = TRUE) / overall_total
}

company_summary <- plastics |>
  filter(!is.na(parent_company),
         parent_company != "Grand Total",
         parent_company != "null",
         parent_company != "NULL",
         parent_company != "Unbranded") |>
  group_by(parent_company) |>
  summarize(
    total_plastic_waste = sum(grand_total, na.rm = TRUE),
    share_of_total = company_share(grand_total, overall_total),
    .groups = "drop"
  ) |>
  arrange(desc(total_plastic_waste)) |>
  slice_head(n = 10)

company_summary
```

### Data Summary 2

**Question**: Which countries had clean up efforts that found the largest grand total of plastics in 2019 and 2020?

```{r}
## Uses iteration and a custom function
total_plastics <- function(input_country){
  
  grand_total_plastics <- plastics |>
    filter(country == input_country) |>
    distinct(year, grand_total) |>
    pull(grand_total)
  
  sum(grand_total_plastics, na.rm = TRUE)
}

plastics |>
  distinct(country) |>
  mutate(
    plastic_total = map_int(country, total_plastics)
    ) |>
  slice_max(plastic_total, n = 10)
```

## Data Summaries That Modify Columns

### Data Summary 3

**Question**: Which countries have the highest average amount of plastic collected per cleanup event?

```{r}
plastic_per_event <- function(input_country) {
  plastics |>
    filter(country == input_country) |>
    distinct(year, grand_total, num_events) |>
    summarise(
      total_plastic = sum(grand_total, na.rm = TRUE),
      total_events = sum(num_events, na.rm = TRUE)
    ) |>
    mutate(plastic_per_event = total_plastic / total_events) |>
    pull(plastic_per_event)
}

plastic_event_summary <- plastics |>
  distinct(country) |>
  mutate(plastic_per_event = map_dbl(country, plastic_per_event)) |>
  slice_max(plastic_per_event, n = 10)

plastic_event_summary

```

### Data Summary 4

**Question**: Which nations had the highest average participation for cleanup events in 2019 and 2020?

```{r}
cleanup_totals <- function(input_country) {
  plastics |>
    filter(country == input_country) |>
    distinct(num_events, volunteers) |>
    mutate(total_vol = sum(volunteers, na.rm = T),
         total_events = sum(num_events, na.rm = T)) |>
    distinct(total_vol, total_events) |>
    mutate(vol_per_event = total_vol / total_events) |>
    pull(vol_per_event)
}

plastics |>
  distinct(country) |>
  mutate(
    avg_participation = map_dbl(country, cleanup_totals),
    rounded_participation = round(avg_participation)
  ) |>
  slice_max(avg_participation, n = 20)
```
