---
title: "Project 1 Draft Report"
format: html
editor: visual
author: "Siddharth Venkiteswaran"
embed-resources: true
warnings: false
code-fold: true
theme: dark
---

## Section 1: Describing the Data

**Plastic Data**

The plastic pollution data was gathered from a publication on Github, that sourced the data from the organization, Break Free From Plastic. Break Free From Plastic is an organization that produces audits on the efforts to clean up plastic pollution within different countries throughout the world. The data set on Github contains data from the 2019 and 2020 audits.

The plastic data is collected with each count of a specific type of plastic, originating from a combination of a specific country and company. This corresponds to a single observational unit in the data. Counts of these plastics are tallied up to create an aggregate of over 50 countries in the data set. The plastics specifically are classified by recycling codes.

The data present here is collected through tallies of clean up efforts across these different countries, per Break Free From Plastic's website. This data was gathered to get information on the sources of plastic pollution, in order to demand a reduction in such waste across the globe, as part of Break Free From Plastic's mission.

From an initial look at what cleaning has been done to the data, it is clear that from the cleaning script, the columns have been set to have values of either values of characters or doubles. This applies for both 2019 and 2020. Furthermore, for 2019, there has been an str_replace operation conducted to better set the names from the file which it was collected from. Additionally, there has been a mutate operation conducted to create new variables of pp and ps, with if else statements to handle NA values within the variable pp_2.

Key information collected:

-   Country of which the plastic was collected in

-   Year of collection (2019 or 2020)

-   Company responsible for creating the plastic waste

-   Grand total of plastics from a country and company combination

**GDP Data**

The GDP data was extracted from API Ninjas' "GDP" API endpoint. Per API Ninjas' website, the data that is provided on their website comes from various different sources, but API Ninjas does not offer much transparency in who specifically gathered the data that is available in the GDP API endpoint. For the GDP endpoint, much of the data goes back to 1980, with the most recent data going up till 2026.

The GDP data is collected with each country being treated as an observational unit. This data follows the nature of a time series, with data being collected back almost 40 to 50 years in the past for each observational unit.

API Ninjas does not offer much transparency in their methods of data collection, only that they gather what is already collected from different sources and compile it together into their API endpoints. Additionally, they do not provide any information on why this data was gathered, only that they aim to provide GDP specific data for comparing one country to another.

During the extraction of the data, the scope was limited to only the years of 2019 and 2020, in accordance with the years that are available from the plastic pollution data.

Key information collected:

-   ISO3 country code

-   Year of observation (2019 or 2020)

-   GDP value (nominal) in US dollars

-   GDP growth rate as a percentage

**Renewable Energy Data**

The renewable energy data was gathered from an open source-data set provided on Github, but it originates from the World Bank's World Development Indicators data set. This data was gathered by the World Bank as part of a project to create a compilation of cross country comparable data on factors regarding the development of countries. The data for many of these indicators goes back for more than 50 years for many of these countries.

The indicators data set is collected with each country being treated as an observational unit. The nature of the data itself is to be treated as time series data that proceeds along until 2022, which is the last year where data was collected in the renewable energy data set.

Per the World Bank's website, much of the data is collected through using official sources reported by these specific countries, with adjustments being made in order to keep the measurements consistent across all countries in the indicators data set. This data was collected in order to provide a better picture of how many countries are faring through their development stages, and to gather information on how to fight poverty in many areas of the world. By having consistent data that can be compared across different areas, the indicators aim to show what differentiates each other from one another.

Some cleaning that was done after the acquisition of the renewable energy data was to filter out all years except for 2019 and 2020, as those were the years pertinent to the plastic data set. Filtering was done using ISO3 codes.

Key information collected:

-   ISO3 country code

-   Year of observation (2019 or 2020)

-   Renewable energy consumption, measured as the percentage of total final energy consumption from renewable sources

## Section 2: Research Questions

There were two main questions that stood out after an initial at all of these data sets.

1.  What companies and countries have the highest number of total plastics thrown out over these two years? And what is the relationship between these countries and the companies responsible for this plastic waste?
2.  Do countries with a higher GDP tend to contribute more to plastic waste? Or is the majority of plastic waste concentrated in mainly developing parts of the world?

## Section 3: Summaries and Visualizations

```{r}
## Installing all of the libraries required for this exploration
library(tidyverse)
library(glue)
library(httr)
library(tidyjson)
library(jsonlite)
library(readr)
library(ggalluvial)
library(ggrepel)
library(paletteer)
library(countrycode)
library(gt)
library(broom)
```

```{r}
## Read in the plastics data
plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')
```

### RQ #1

What companies and countries have the highest number of total plastics thrown out over these two years? And what is the relationship between these countries and the companies responsible for this plastic waste?

```{r}
company_total <- function(input_company) {
  
  company_grand_total <- plastics |>
    filter(parent_company == input_company) |>
    distinct(year, grand_total) |>
    pull(grand_total) 
  
  sum(company_grand_total, na.rm = TRUE)
  
}

plastics |>
  distinct(parent_company) |>
  filter(!is.na(parent_company),
         parent_company != "Grand Total",
         parent_company != "null",
         parent_company != "NULL",
         parent_company != "Unbranded") |>
  mutate(
    total_plastic_waste = map_int(parent_company, company_total),
    share_of_total = total_plastic_waste / sum(total_plastic_waste, na.rm = TRUE)
  ) |>
  slice_max(total_plastic_waste, n = 10) |>
  gt() |>
  tab_header(
    title = "Company Contribution to Plastic Waste"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_title()
  ) |>
  cols_label(
    parent_company = "Parent Company",
    total_plastic_waste = "Total Waste",
    share_of_total = "Share of Total"
  ) |>
  fmt_percent(
    columns = share_of_total,
    decimals = 3
  ) |>
  tab_caption("Table 1: Amount of Plastic Waste Attributed to Each Company") |>
  tab_spanner(
    label = "Plastic Counts",
    columns = c(total_plastic_waste, share_of_total)
  )

```

### Data Summary 2

**Question**: Which countries had clean up efforts that found the largest grand total of plastics in 2019 and 2020?

```{r}
## Uses iteration and a custom function
total_plastics <- function(input_country){
  
  grand_total_plastics <- plastics |>
    filter(country == input_country) |>
    distinct(year, grand_total) |>
    pull(grand_total)
  
  sum(grand_total_plastics, na.rm = TRUE)
}

plastics |>
  distinct(country) |>
  filter(country != "EMPTY") |>
  mutate(
    plastic_total = map_int(country, total_plastics)
    ) |>
  slice_max(plastic_total, n = 10)
```

## Data Summaries That Modify Columns

### Data Summary 3

**Question**: Which countries have the highest average amount of plastic collected per cleanup event?

```{r}
plastic_per_event <- function(input_country) {
  plastics |>
    filter(country == input_country) |>
    distinct(year, grand_total, num_events) |>
    summarise(
      total_plastic = sum(grand_total, na.rm = TRUE),
      total_events = sum(num_events, na.rm = TRUE)
    ) |>
    mutate(plastic_per_event = total_plastic / total_events) |>
    pull(plastic_per_event)
}

plastic_event_summary <- plastics |>
  distinct(country) |>
  mutate(plastic_per_event = map_dbl(country, plastic_per_event)) |>
  slice_max(plastic_per_event, n = 10)

plastic_event_summary

```

### Data Summary 4

**Question**: Which nations had the highest average participation for cleanup events in 2019 and 2020?

```{r}
cleanup_totals <- function(input_country) {
  plastics |>
    filter(country == input_country) |>
    distinct(num_events, volunteers) |>
    mutate(total_vol = sum(volunteers, na.rm = T),
         total_events = sum(num_events, na.rm = T)) |>
    distinct(total_vol, total_events) |>
    mutate(vol_per_event = total_vol / total_events) |>
    pull(vol_per_event)
}

plastics |>
  distinct(country) |>
  mutate(
    avg_participation = map_dbl(country, cleanup_totals),
    rounded_participation = round(avg_participation)
  ) |>
  slice_max(avg_participation, n = 20)
```

## Project Checkpoint #3

**Create a function to pull data from the API**

```{r}
#| eval: false
get_gdp <- function(country_code, key = my_key) {
  url <- glue("https://api.api-ninjas.com/v1/gdp?country={country_code};X-Api-Key={key}")
  
  my_data <- safely(fromJSON)(url)
  
  if (is.null(my_data$result)) {
    return(NA)
  } else {
    return(my_data$result |> filter(year %in% c(2019, 2020)))
  }
}
```

**Obtain a list of the top 10 countries in plastic waste totals**

```{r}
## Create a list of the top 10 countries with the highest total plastics


top_plastic_countries <- plastics |>
  distinct(country) |>
  filter(!country %in% c("EMPTY", "Taiwan_ Republic of China (ROC)")) |>  ## Dropping Taiwan because of its contested status in the world (some supplemental data does not include Taiwan)
  mutate(
    plastic_total = map_int(country, total_plastics),
    country_code = countrycode(country, origin = "country.name", destination = "iso3c")
  ) |>
  filter(!is.na(country_code)) |> ## Dropping Nigeria's duplicate
  filter(plastic_total != 61123) |> ## Added security for removing the specific duplicate
  slice_max(plastic_total, n = 50) |>
  pull(country_code)

```

**Pull data from the API**

```{r}
#| eval: false
Sys.sleep(20)
## Create a list of the countries with their GDP
gdp_list <- top_plastic_countries |>
  map(get_gdp)

## Double check for NAs
gdp_list |>
  is.na() |>
  sum()
```

**Turn the data into a CSV file**

```{r}
#| eval: false
## Turn this into a tibble to join with our original data frame
gdp_data <- gdp_list |>
  keep(is.data.frame) |>
  bind_rows() |>
  rename(country_code = country)

gdp_data |>
  pull(country_code)

## Saving the csv locally
write.csv(gdp_data, "gdp.csv", row.names = FALSE)
```

```{r}
## Rereading in the data from the local file
gdp_df <- read.csv("gdp.csv")
```

**Join the data sets together**

```{r}
#| warning: false
plastics_top <- plastics |>
  filter(!country %in% c("EMPTY", 
                         "Nigeria", 
                         "Taiwan_ Republic of China (ROC)", 
                         "United Kingdom")) |>
  mutate(
    country_code = countrycode(country, origin = "country.name", destination = "iso3c")
  ) |>
  filter(country_code %in% top_plastic_countries) |>
  left_join(gdp_df)

head(plastics_top)
```

```{r}
renewable_data <- read.csv("renewable-energy.csv")
## Data found from https://github.com/datasets/world-development-indicators/tree/main/indicators/eg.fec.rnew.zs

plastics_top <- plastics_top |>
  left_join(renewable_data |>
              filter(Year %in% c(2019, 2020),
                     Country.Code %in% plastics_top$country_code), 
            by = c("country_code" = "Country.Code", 
                   "year" = "Year")) |>
  select(-Country.Name)
```

### Additional Data Summaries/Visualizations

**RQ:** What is the relationship between total plastic was and percentage of renewable energy, for the top 10 countries in total plastic waste ?

```{r}
renewable_summary <- plastics_top |>
  filter(country_code %in% top_plastic_countries) |>
  distinct(country, 
           country_code, 
           year, 
           grand_total, 
           Value) |>
  group_by(country, 
           country_code) |>
  summarize(
    total_waste = sum(grand_total, na.rm = TRUE),
    renewable_pct = mean(Value, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(country != "Ecuador") |>
  arrange(desc(total_waste))

renewable_summary
```

**New Plot:** Plastic Waste vs. Renewable Energy Percentage by Country

```{r}
ggplot(renewable_summary, aes(x = total_waste, y = renewable_pct)) +
  geom_point(size = 3, color = "black") +
  geom_text_repel(
    aes(label = country_code),
    size = 4,
    box.padding = 0.5,
    point.padding = 0.5
  ) +
  labs(
    title = "Plastic Waste vs Renewable Energy by Country",
    subtitle = "Top 50 Countries by Total Plastic Waste",
    x = "Total Plastic Waste",
    y = "Renewable Energy (% of total energy consumption)"
  ) +
  theme_classic(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    plot.margin = margin(15, 15, 15, 15),
    axis.title = element_text(size = 12)
  )

```

**New Plot:** GDP vs. Total Plastic Waste by Country

```{r}

```

```{r}
top10_codes <- plastics_top |>
  distinct(country_code) |>
  slice_head(n = 10) |>
  pull(country_code)

country_plot <- plastics_top |>
  filter(country_code %in% top10_codes) |>
  group_by(country, country_code) |>
  summarize(
    total_plastic = sum(grand_total, na.rm = TRUE),
    gdp = mean(gdp_nominal, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(country_plot, aes(x = gdp, y = total_plastic)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_text(aes(label = country_code), vjust = -0.5, size = 3) +
  labs(
    title = "Relationship Between GDP and Total Plastics (Top 10 Countries)",
    x = "GDP (billions USD)",
    y = "Total Plastics"
  ) +
  theme_minimal()

```

## Project Checkpoint #4

**Improved Plot**

```{r}
#| fig-alt: "Scatterplot of total plastic waste versus renewable energy consumption by country, with most countries clustered at low waste levels and a few high-waste outliers such as Nigeria. The plot shows no strong relationship between the variables."

renewable_summary <- renewable_summary |>
  mutate(
    group = case_when(
      country_code %in% c("NGA", "CHE", "PHL") ~ country_code,
      total_waste < 40000 & renewable_pct < 40 ~ "Low waste / Low renewable",
      total_waste < 40000 & renewable_pct >= 40 ~ "Low waste / High renewable",
      TRUE ~ "Other"
    ), 
    group = factor(group, 
      levels = c("Low waste / Low renewable",
                 "Low waste / High renewable",
                 "NGA",
                 "PHL",
                 "CHE")
    )
  )

ggplot(renewable_summary, aes(x = total_waste, y = renewable_pct, color = group)) +
  geom_point(size = 2) +
  geom_text_repel(
    aes(label = country_code),
    size = 2
  ) +
  labs(
    title = "Plastic Waste vs Renewable Energy by Country",
    subtitle = "Top 50 Countries by Total Plastic Waste",
    x = "Total Plastic Waste",
    y = "Renewable Energy (% of total energy consumption)",
    color = "Group"
  ) +
  scale_color_manual(values = c(
  "Low waste / Low renewable" = "#1f77b4",  
  "Low waste / High renewable" = "#2ca02c", 
  "Other" = "gray70",
  "NGA" = "#e41a1c", 
  "CHE" = "#ff7f00",  
  "PHL" = "#000000"    
)) +
  theme_classic() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    plot.margin = margin(15, 15, 15, 15))
```

**RQ:** Which countries have the highest amount of plastic waste, and what from what companies does that waste originate from?

```{r}
#| fig-alt: "Sankey diagram showing how plastic waste flows between countries and companies. The diagram is stratified by the type of plastic. The Philippines and Nigeria have the highest amount of waste flowing into a variety of different companies. The most common type of plastic waste is PET."

## Get a list of the top 10 companies and the top 5 countries in plastic waste
top10_companies <- unique(company_summary$parent_company)
top5_codes <- top_plastic_countries[1:5]

## Convert the data into the long format
plastics_long <- plastics_top |>
  filter(parent_company %in% top10_companies,
         country_code %in% top5_codes) |>
  select(country, year, parent_company, hdpe, ldpe, o, pet, pp, ps, pvc) |>
  pivot_longer(
    cols = c(hdpe, ldpe, o, pet, pp, ps, pvc),
    names_to = "plastic_type",
    values_to = "total_plastic"
  ) |>
  filter(total_plastic > 0)

## Create the Sankey diagram
plastics_long |>
  mutate(
    country = fct_reorder(country, 
                          total_plastic, 
                          sum, 
                          .desc = TRUE),
    parent_company = fct_reorder(parent_company, 
                                 total_plastic, 
                                 sum, 
                                 .desc = TRUE),
    parent_company = recode(parent_company,
    "Universal Robina Corporation" = "Universal Robina",
    "The Coca-Cola Company" = "Coca-Cola")
  ) |> 
  ggplot(aes(axis1 = country,
             axis2 = parent_company,
             y = total_plastic)) +
  scale_x_discrete(limits = c("Country", "Company"), 
                   expand = c(.2, .05)) +
  geom_alluvium(aes(fill = plastic_type)) +
  geom_stratum() +
  ggfittext::geom_fit_text(stat = "stratum", 
                           aes(label = after_stat(stratum)),  
                           width = 1/4, 
                           min.size = 3) +
  theme_bw() +
  scale_fill_paletteer_d("ggthemes::Color_Blind") +
  xlab("Countries vs. Companies") +
  ylab("Total Plastic") +
  ggtitle("Plastic Waste from Country to Company",
          "Stratified by Plastic Type") + 
  labs(fill = "Plastic Type")
```

A simple linear regression model was used:

$Y_i$ = $\beta_0$ + $\beta_1 X_i$ + $\varepsilon_i$

$Y_i$ = renewable energy percentage for country $i$

$X_i$ = total plastic waste for country $i$

$\beta_0$ = intercept

$\beta_1$ = slope parameter

$\varepsilon_i\sim \text{iid } N(0, \sigma^2)$

```{r}
## Used the broom package to better format our summary https://broom.tidymodels.org/
model <- lm(renewable_pct ~ total_waste, data = renewable_summary)

tidy(model) |>
  gt() |>
  tab_header(
    title    = "Linear Regression Results",
    subtitle = "Response: Renewable Energy Percentage | Predictor: Total Waste"
  ) |>
  tab_style(style = cell_text(weight = "bold"),
    locations = list(cells_column_labels(), 
                     cells_title())) |>
  tab_caption("Table 1: Effect of Total Waste on Renewable Energy Percentage") |>
  cols_label(
    term      = "Term",
    estimate  = "Estimate",
    std.error = "Std. Error",
    statistic = "t-Statistic",
    p.value   = "p-Value"
  ) |>
  fmt_number(
    columns  = c(estimate, std.error, statistic),
    decimals = 4
  ) |>
  fmt_number(
    columns  = p.value,
    decimals = 4
  ) 

```

A simple linear regression model was conducted to explore the relationship between total plastic waste and renewable energy percentage among the top 10 countries. The estimated slope was 0.0002, showing a slight positive relationship. However, this relationship was not statistically significant from the p value of 0.0905, leading to the conclusion that there is insufficient evidence to conclude that total plastic waste is associated with renewable energy use in these countries.
