---
title: "Waste and Pollution in an Era of Excess"
subtitle: "Where does plastic waste originate from and where is it concentrated in?"
format: html
editor: visual
author: "Siddharth Venkiteswaran"
embed-resources: true
warnings: false
code-fold: true
theme: cosmo
---

## Why is this important?

Plastic waste pollution has been a problem that has plagued the world for some time, with the magnitude of the problem continuing to grow over the years as efforts to keep waste levels in check have mostly been minimal at best. The amount of waste continues to be a looming threat to the health of many different ecosystems, and this will only continue to get worse for future generations. To better understand this problem, we must better examine the sources of where all this pollution comes from, and understand what exactly rising levels of plastic waste are correlated with. By doing so, we can leverage these relationships to better address the problem of rising levels of waste, and provide a more stable setup for cleanup efforts into the future.

The main item that motivated this data exploration was the relationship of plastic waste with different countries across the globe. Different parts of the globe are subject to different situations, so evidently their relationship with the problem of plastic waste would be different. Two questions came up that appeared to warrant a deeper investigation.

The first question was in regards to the relationship between a country and a corporation when it came to plastic waste. What companies and countries have the highest number of total plastics thrown out over these two years? And were the top countries in plastic waste tied with specific companies that managed to produce products that were responsible for such high levels of pollution?

The second question was more specifically directed towards the countries around the globe. Different countries reside in different levels of economic status, some are richer than others and have a larger footprint in the global economy. Do these richer countries tend to contribute more to plastic waste levels due to their larger economic footprint? Or is the majority of plastic waste concentrated within areas of the world that are considered to be developing regions? One would assume that a country that has a large production output, would also have a large waste output too, and the idea of figuring out if this was true or not was one that motivated this data exploration.

## Understanding the data sets at hand

There were two data sets that were chosen to carry out this investigation.

The first data set comes from a publication on Github, which sourced data from the organization, Break Free From Plastic. Break Free From Plastic is an organization that produces audits on the efforts to clean up plastic pollution within different countries throughout the world. The data set on Github contains data from the 2019 and 2020 audits. With data on different countries around the world in regards to plastic waste, this appeared to be a data set that was relevant to the questions that motivated this exploration.

The plastic data was collected with each count of a specific type of plastic, originating from a combination of a specific country and company. This corresponded to a single observational unit in the data. Counts of these plastics were tallied up to create an aggregate of over 50 countries in the data set. The plastics specifically were classified by recycling codes.

From Break Free From Plastic's website, the data that was present in the Github publication was collected through tallies of clean up efforts across these different countries. This data was gathered to get information on the sources of plastic pollution, in order to demand a reduction in such waste across the globe, as part of Break Free From Plastic's mission.

The Github publication also contained a cleaning script on what was done to the data before it was published. From the script, the main cleaning operations that were done were:

1.  Setting the columns to have values of either characters or doubles for both 2019 and 2020
2.  Conducting a str_replace operation to better set the names from the file which it was collected from
3.  Conducting a mutate operation to create new variables of pp and ps, with if else statements to handle NA values within the variable pp_2

Overall the main variables from this data set were:

- Country of which the plastic was collected in

- Year of collection (2019 or 2020)

- Company responsible for creating the plastic waste

- Grand total of plastics from a country and company combination

The second data set was a GDP data set in regards to specific countries, that was published by the website API Ninjas. Per API Ninjas' website, the data that is provided on their website comes from various different sources, but API Ninjas does not offer much transparency in who specifically gathered the data that is available in the GDP API endpoint. For the GDP endpoint, much of the data goes back to 1980, with the most recent data going up till 2026.

Data was collected with each country being treated as an observational unit. This data follows the nature of a time series, with data being collected back almost 40 to 50 years in the past for each observational unit.

API Ninjas does not offer much transparency in their methods of data collection, only that they gather what is already collected from different sources and compile it together into their API endpoints. Additionally, from the website, there was no information provided on why this data was gathered, only that the data was collected with the aim of comparing GDP data from one country with others.

The key information that was present in this data set was:

- ISO3 country code

- Year of observation (2019 or 2020)

- GDP value per capita (nominal) in US dollars

- GDP growth rate as a percentage

With these data sets present, the data exploration began by starting off with looking at the relationship between a country and a corporation in regards to plastic waste, before carrying out a country level analysis on how the GDP per capita of a country corresponded to plastic waste.

```{r}
#| output: false
library(tidyverse)
library(glue)
library(httr)
library(tidyjson)
library(jsonlite)
library(readr)
library(ggalluvial)
library(ggrepel)
library(paletteer)
library(countrycode)
library(gt)
library(broom)
```

```{r}
#| output: false
## Read in the plastics data
plastics <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2021/2021-01-26/plastics.csv')
```

```{r}
## Custom function to get the total number of plastics for each country
total_plastics <- function(input_country){
  
  grand_total_plastics <- plastics |>
    filter(country == input_country) |>
    distinct(year, grand_total) |>
    pull(grand_total)
  
  sum(grand_total_plastics, na.rm = TRUE)
}

## Custom function to get total number of plastics for each company
company_total <- function(input_company) {
  
  company_grand_total <- plastics |>
    filter(parent_company == input_company) |>
    distinct(year, grand_total) |>
    pull(grand_total) 
  
  sum(company_grand_total, na.rm = TRUE)
  
}

## Custom function for outputting average gdp and plastic total
gdp_plastics <- function(input_country){
  
  country_data <- plastics_top |>
    filter(country == input_country) |>
    distinct(year, grand_total, gdp_per_capita_nominal)
  
  tibble(
    plastic_total = sum(country_data$grand_total, na.rm = TRUE),
    avg_gdp_per_capita_nominal = mean(country_data$gdp_per_capita_nominal, na.rm = TRUE)
  )
}

```

```{r}
#| eval: false
get_gdp <- function(country_code, key = my_key) {
  url <- glue("https://api.api-ninjas.com/v1/gdp?country={country_code};X-Api-Key={key}")
  
  my_data <- safely(fromJSON)(url)
  
  if (is.null(my_data$result)) {
    return(NA)
  } else {
    return(my_data$result |> filter(year %in% c(2019, 2020)))
  }
}
```

```{r}
## Create a list of the top 10 countries with the highest total plastics
top_plastic_countries <- plastics |>
  distinct(country) |>
  filter(!country %in% c("EMPTY", "Taiwan_ Republic of China (ROC)")) |>  ## Dropping Taiwan because of its contested status in the world (some supplemental data does not include Taiwan)
  mutate(
    plastic_total = map_int(country, total_plastics),
    country_code = countrycode(country, origin = "country.name", destination = "iso3c")
  ) |>
  filter(!is.na(country_code)) |> ## Dropping Nigeria's duplicate
  filter(plastic_total != 61123) |> ## Added security for removing the specific duplicate
  slice_max(plastic_total, n = 50) |>
  pull(country_code)
```

```{r}
#| eval: false
Sys.sleep(20)
## Create a list of the countries with their GDP
gdp_list <- top_plastic_countries |>
  map(get_gdp)

```

```{r}
#| eval: false
## Turn this into a tibble to join with our original data frame
gdp_data <- gdp_list |>
  keep(is.data.frame) |>
  bind_rows() |>
  rename(country_code = country)

gdp_data |>
  pull(country_code)

## Saving the csv locally
write.csv(gdp_data, "gdp.csv", row.names = FALSE)
```

```{r}
## Rereading in the data from the local file
gdp_df <- read.csv("gdp.csv")
```

```{r}
#| warning: false
plastics_top <- plastics |>
  filter(!country %in% c("EMPTY", 
                         "Nigeria", 
                         "Taiwan_ Republic of China (ROC)", 
                         "United Kingdom")) |>
  mutate(
    country_code = countrycode(country, origin = "country.name", destination = "iso3c")
  ) |>
  filter(country_code %in% top_plastic_countries) |>
  left_join(gdp_df)
```

## Coca Cola is responsible for a large amount of waste

```{r}
company_summary <- plastics |>
  distinct(parent_company) |>
  filter(!is.na(parent_company),
         parent_company != "Grand Total",
         parent_company != "null",
         parent_company != "NULL",
         parent_company != "Unbranded") |>
  mutate(
    total_plastic_waste = map_int(parent_company, company_total),
    share_of_total = total_plastic_waste / sum(total_plastic_waste, na.rm = TRUE)
  ) |>
  slice_max(total_plastic_waste, n = 10) 
  
company_summary |>   
  gt() |>
  tab_header(
    title = "Company Contribution to Plastic Waste"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_title()
  ) |>
  cols_label(
    parent_company = "Parent Company",
    total_plastic_waste = "Total Waste",
    share_of_total = "Share of Total"
  ) |>
  fmt_percent(
    columns = share_of_total,
    decimals = 3
  ) |>
  fmt_number(columns = total_plastic_waste,
             decimals = 0) |>
  tab_caption("Table 1: Amount of Plastic Waste Attributed to Each Company") |>
  tab_spanner(
    label = "Plastic Counts",
    columns = c(total_plastic_waste, share_of_total)
  )

```

The Coca-Cola Company leads the counts of plastic waste by a comfortable margin, having over 10,000 more items of plastic waste compared to the company La Doo, which has a clear grip on second place. After those two companies, the amount of plastic waste is very close among these different companies, indicating that perhaps the top two companies could be outliers in comparison to the other corporations.

## Two countries carry extreme amounts of plastic waste

```{r}
## Uses iteration and a custom function
plastics |>
  distinct(country) |>
  filter(!country %in% c("EMPTY", "Taiwan_ Republic of China (ROC)", "Nigeria")) |>
  mutate(
    plastic_total = map_int(country, total_plastics),
    country = recode(country, "NIGERIA" = "Nigeria", "ECUADOR" = "Ecuador"),
    share_of_total = plastic_total / sum(plastic_total, na.rm = TRUE)
    ) |>
  slice_max(plastic_total, n = 10) |>
  gt() |>
  tab_header(
    title = "Countries with the Most Plastic Waste"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_title()
  ) |>
  cols_label(
    country = "Country",
    plastic_total = "Total Waste",
    share_of_total = "Share of Total"
  ) |>
  tab_caption("Table 2: Amount of Plastic Waste From Each Country") |>
  tab_footnote(
    footnote = "Taiwan and Republic of China were removed from data set due to the contested nature of the region. Additionally, duplicates were removed by choosing to keep the entry with higher values for plastic waste.",
    locations = cells_column_labels(
      columns = country
    )
  ) |>
  fmt_number(columns = plastic_total,
             decimals = 0) |>
  fmt_percent(
    columns = share_of_total,
    decimals = 3
  ) |>
  tab_spanner(
    label = "Plastic Counts",
    columns = c(plastic_total, share_of_total)
  )

```

The top country in terms of plastic waste is Nigeria, followed by the Philippines. In a similar trend to what was apparent from Table 1, the top two in this table seem to be outliers in comparison to the rest of the world, as they are the only countries with plastic waste counts above 100,000. The rest of the world seems to be more closely clustered with each other in terms of amount of plastic waste.

## Which companies and countries have the most waste?

```{r}
#| fig-alt: "Sankey diagram showing how plastic waste flows between countries and companies. The diagram is stratified by the type of plastic. The Philippines and Nigeria have the highest amount of waste flowing into a variety of different companies. The most common type of plastic waste is PET."

## Get a list of the top 10 companies and the top 5 countries in plastic waste
top10_companies <- unique(company_summary$parent_company)
top5_codes <- top_plastic_countries[1:5]

## Convert the data into the long format
plastics_long <- plastics_top |>
  filter(parent_company %in% top10_companies,
         country_code %in% top5_codes) |>
  select(country, year, parent_company, hdpe, ldpe, o, pet, pp, ps, pvc) |>
  pivot_longer(
    cols = c(hdpe, ldpe, o, pet, pp, ps, pvc),
    names_to = "plastic_type",
    values_to = "total_plastic"
  ) |>
  filter(total_plastic > 0)

## Create the Sankey diagram
plastics_long |>
  mutate(
    country = fct_reorder(country, 
                          total_plastic, 
                          sum, 
                          .desc = TRUE),
    parent_company = fct_reorder(parent_company, 
                                 total_plastic, 
                                 sum, 
                                 .desc = TRUE),
    parent_company = recode(parent_company,
    "Universal Robina Corporation" = "Universal Robina",
    "The Coca-Cola Company" = "Coca-Cola"),
    country = recode(country,
                     "NIGERIA" = "Nigeria")
  ) |> 
  ggplot(aes(axis1 = country,
             axis2 = parent_company,
             y = total_plastic)) +
  scale_x_discrete(limits = c("Country", "Company"), 
                   expand = c(.2, .05)) +
  geom_alluvium(aes(fill = plastic_type)) +
  geom_stratum() +
  ggfittext::geom_fit_text(stat = "stratum", 
                           aes(label = after_stat(stratum)),  
                           width = 1/4, 
                           min.size = 3) +
  theme_bw() +
  scale_fill_paletteer_d("ggthemes::Color_Blind") +
  xlab("Countries vs. Companies") +
  ylab("Total Plastic") +
  ggtitle("Plastic Waste from Country to Company",
          "Stratified by Plastic Type") + 
  labs(fill = "Plastic Type")
```

The Sankey diagram above shows the relationship between a country and a company. The countries were limited to the top 5 countries in plastic waste so that the plot could be more easily viewed, while the number of companies is limited to the top 10. From the plot, it is clear that Nigeria is closely tied to La Doo, while the Philippines is more spread out in terms of which corporations contribute to its plastic waste. Additionally, while Coca Cola had the greatest overall total of plastic waste out of all of the companies, within the top 5 countries in plastic waste, it ranks a bit lower, which indicates that the plastic waste that it is responsible for is spread out over a much larger set of countries, rather than just being concentrated in one.

## Plastic waste seems to be spread out

```{r}
#| fig-alt: "Heatmap showing where the most dense concentration of plastic waste lies between combinations of country and parent company. From the plot, the densest concentration of plastic waste lies in the company of La Doo and Nigeria. Additionally, the Coca-Cola Company has its plastic waste spread out across most of the top 5 countries in plastic waste, with the most waste being in the Philippines."

heatmap_data <- plastics_top |>
  filter(
    !is.na(parent_company),
    !country %in% c("EMPTY", "Taiwan_ Republic of China (ROC)", "Nigeria"),
    parent_company != "Grand Total",
    parent_company != "null",
    parent_company != "NULL",
    parent_company != "Unbranded"
  ) |>
  group_by(country, parent_company, country_code) |>
  summarize(plastic_total = sum(grand_total, na.rm = TRUE), .groups = "drop")

heatmap_data |>
  filter(country_code %in% top5_codes, parent_company %in% top10_companies) |>
  mutate(country = recode(country,
                          "NIGERIA" = "Nigeria")) |>
  ggplot(aes(x = country, y = parent_company, fill = plastic_total)) +
  geom_tile() +
  labs(title = "Density of Plastic Waste Between Countries and Companies",
       x = "Country",
       y = "Parent Company",
       fill = "Total Plastic") +
  theme_bw() +
  scale_fill_viridis_c() + 
  theme(plot.title = element_text(hjust = 0.5))
```

The heat map above complements the findings from the Sankey diagram, as it shows again how Nigeria is very closely tied with the company of La Doo, while showing the rest of the corporations do not hold any such relationship with any other countries. The only country that seems to have a close relationship with a corporation, like Nigeria does with La Doo, is the Philippines with the Universal Robina Corporation. But even with that relationship, the Philippines plastic waste is influenced by a variety of different countries, not just one main one.

From both diagrams, it would appear that there does not appear to be any close relationships between countries and companies when it comes to the amount of plastic waste that is attributed to each of them. The top companies in regards to amount of plastic waste in the world get to that status by spreading the wealth between a variety of different countries, rather than having a close relationship with one. This would indicate that the corporations most responsible for large amounts of plastic waste have a global footprint, rather than a local one.

## How rich are countries in the top 10 of plastic waste?

```{r}
plastics_top |>
  distinct(country) |>
  filter(!country %in% c("EMPTY", "Taiwan_ Republic of China (ROC)", "Nigeria")) |>
  mutate(
    result = map(country, gdp_plastics),
    country = recode(country, "NIGERIA" = "Nigeria", "ECUADOR" = "Ecuador")
  ) |>
  unnest(result) |>
  slice_max(plastic_total, n = 10) |>
  gt() |>
  tab_header(
    title = "Comparing Average GDP with Plastic Waste"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_title()
  ) |>
  cols_label(
    country = "Country",
    plastic_total = "Total Waste",
    avg_gdp_per_capita_nominal = "GDP Per Capita (Average)"
  ) |>
  tab_caption("Table 3: Average GDP Compared to Plastic Waste From Each Country") |>
  tab_spanner(
    label = "Country Statistics",
    columns = c(plastic_total, avg_gdp_per_capita_nominal)
  ) |>
  fmt_currency(columns = avg_gdp_per_capita_nominal,
             decimals = 2) |>
  tab_footnote(
    footnote = "Expressed in billions of USD",
    locations = cells_column_labels(
      columns = avg_gdp_per_capita_nominal
    )) |>
  fmt_number(columns = plastic_total,
             decimals = 0)
```

From Table 3, most of the countries in the top 10 of plastic waste have low average GDP per capita numbers, with the United States and Switzerland being the main outliers here. This would initially disprove the intuitive belief that countries with a higher gross domestic output, meaning they produce more, do not contribute high levels of waste compared to countries with a lower average GDP per capita.

## Countries with low GDP per capita have more waste

```{r}
## Modify the plastics data to get 3 quantiles and classifications
quantile_data <- plastics_top |>
  distinct(country, year, grand_total, gdp_per_capita_nominal) |>
  mutate(
    gdp_quantile = case_when(
      gdp_per_capita_nominal <= quantile(gdp_per_capita_nominal, 
                                         0.33, 
                                         na.rm = TRUE) ~ "Low GDP",
      gdp_per_capita_nominal <= quantile(gdp_per_capita_nominal, 
                                         0.66, 
                                         na.rm = TRUE) ~ "Mid GDP",
      TRUE ~ "High GDP"
    ),
    gdp_quantile = factor(gdp_quantile,
                          levels = c("Low GDP", "Mid GDP", "High GDP"))
  )

## Create functions to get the total plastic waste and the average gdp per capita
quantile_plastic <- function(input_quantile) {
  
  total <- quantile_data |>
    filter(gdp_quantile == input_quantile) |>
    pull(grand_total)
  
  sum(total, na.rm = TRUE)
}
 
quantile_gdp <- function(input_quantile) {
  
  avg_gdp_by_country <- quantile_data |>
    filter(gdp_quantile == input_quantile) |>
    group_by(country) |>
    summarise(avg_gdp = mean(gdp_per_capita_nominal, na.rm = TRUE), .groups = "drop") |>
    pull(avg_gdp)
  
  mean(avg_gdp_by_country, na.rm = TRUE)
}

## Iterate and modify the table
tibble(gdp_quantile = c("Low GDP", "Mid GDP", "High GDP")) |>
  mutate(
    total_plastic = map_int(gdp_quantile, quantile_plastic),
    avg_gdp_per_capita_nominal = map_dbl(gdp_quantile, quantile_gdp),
    share_of_total  = total_plastic / sum(total_plastic, na.rm = TRUE)
  ) |>
  gt() |>
  tab_header(
    title = "Plastic Waste by GDP Categories"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_title()
  ) |>
  cols_label(
    gdp_quantile = "GDP Category",
    total_plastic = "Total Plastic",
    avg_gdp_per_capita_nominal = "GDP Per Capita (Average)",
    share_of_total = "Share of Total"
  ) |>
  tab_caption("Table 4: Plastic Waste Attributed to Different GDP Levels") |>
  tab_spanner(
    label = "GDP Statistics",
    columns = c(avg_gdp_per_capita_nominal, share_of_total)) |>
  tab_footnote(
    footnote = "Expressed in billions of USD",
    locations = cells_column_labels(
      columns = avg_gdp_per_capita_nominal
    )) |>
  fmt_currency(columns = avg_gdp_per_capita_nominal,
             decimals = 2) |>
  fmt_number(columns = total_plastic,
             decimals = 0) |>
  fmt_percent(columns = share_of_total,
              decimals = 3)


```

Table 4 adds more support to the belief that countries with lower GDP per capita have larger amounts of plastic waste. Table 4's groupings of all the countries into different quantiles corresponding with average GDP per capita show that countries with low GDP have way higher amounts of plastic waste compared to mid and high GDP countries. In fact, there is appears to be a negative relationship present, meaning as GDP status increases, the amount of plastic decreases. However, caution must be exercised here, as the low GDP category contains two major outliers in terms of plastic waste total, in Nigeria and the Philippines. Further analysis must be conducted to see if there really is a relationship present here.

## Comparing rankings

```{r}
#| fig-alt: "Dumbbell plot comparing GDP per capita rank and plastic waste rank across countries. Blue points represent GDP per capita rank, orange points represent plastic waste rank, and gray lines show the gap between them."

rq2_dumbbell <- plastics_top %>%
  filter(!is.na(gdp_per_capita_nominal), !is.na(grand_total)) %>%
  group_by(country) %>%
  summarize(
    total_plastic = sum(grand_total, na.rm = TRUE),
    avg_gdp_per_capita = mean(gdp_per_capita_nominal, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    gdp_rank = min_rank(desc(avg_gdp_per_capita)),
    plastic_rank = min_rank(desc(total_plastic)),
    rank_gap = abs(gdp_rank - plastic_rank)
  ) %>%
  slice_max(total_plastic, n = 10) %>%
  mutate(country = fct_reorder(country, rank_gap))

ggplot(rq2_dumbbell) +
  geom_segment(
    aes(
      y = country, yend = country,
      x = gdp_rank, xend = plastic_rank
    ),
    color = "gray60",
    linewidth = 1.3
  ) +
  geom_point(
    aes(x = gdp_rank, y = country, color = "GDP per Capita Rank"),
    size = 3
  ) +
  geom_point(
    aes(x = plastic_rank, y = country, color = "Plastic Waste Rank"),
    size = 3
  ) +
  scale_color_manual(
    name = "Legend",
    values = c(
      "GDP per Capita Rank" = "steelblue",
      "Plastic Waste Rank" = "darkorange"
    )
  ) +
  labs(
    title = "GDP per Capita Rank vs Plastic Waste Rank by Country",
    subtitle = "Gray lines show the gap between GDP per capita rank (blue) and plastic waste rank (orange)",
    x = "Rank (1 = highest)",
    y = "Country"
  ) +
  theme_minimal(base_size = 12)
```

```{r}
#| fig-alt: "A scatterplot of total plastic waste vs nominal GDP colored by income group. Lower middle income countries like Nigeria and the Philippines produce the most plastic waste despite low GDPs, while high income countries show moderate waste levels despite higher GDPs, suggesting no clear positive relationship between GDP and plastic waste."
#| warning: false

full_summary <- plastics_top |>
  distinct(country) |>
  filter(!country %in% c("EMPTY", "Taiwan_ Republic of China (ROC)", "Nigeria")) |>
  mutate(
    result = map(country, gdp_plastics),
    country = recode(country, "NIGERIA" = "Nigeria", "ECUADOR" = "Ecuador")
  ) |>
  unnest(result) |>
  mutate(gdp_category = case_when(
    is.na(avg_gdp_per_capita_nominal) ~ "Unknown",
    avg_gdp_per_capita_nominal < 5000  ~ "Low GDP",
    avg_gdp_per_capita_nominal < 20000 ~ "Middle GDP",
    avg_gdp_per_capita_nominal >= 20000 ~ "High GDP"
  )) 

full_summary |>
  filter(country != "China") |>
  ggplot(aes(x = avg_gdp_per_capita_nominal, y = plastic_total, color = gdp_category)) +
    geom_point(size = 1.5) +
    geom_text_repel(
      aes(label = country),
      size = 2
  ) +
  labs(
    title    = "Total Plastic Waste vs GDP by Country",
    subtitle = "Colored by GDP Category",
    x        = "Avg GDP Per Capita(Nominal, USD)",
    y        = "Total Plastic Waste",
    color    = "GDP Category"
  ) +
  scale_color_manual(values = c(
    "Low GDP"          = "#e41a1c",
    "Middle GDP" = "#BF00FF",
    "High GDP"         = "#1f77b4"
  )) +
  theme_classic() +
  theme(
    plot.title    = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12),
    plot.margin   = margin(15, 15, 15, 15)
  )
```

A simple linear regression model was used:

$Y_i$ = $\beta_0$ + $\beta_1 X_i$ + $\varepsilon_i$

$Y_i$ = total plastic waste for country $i$

$X_i$ = GDP per capita (nominal) for country $i$

$\beta_0$ = intercept

$\beta_1$ = slope parameter

$\varepsilon_i\sim \text{iid } N(0, \sigma^2)$

```{r}
model_gdp <- lm(plastic_total ~ avg_gdp_per_capita_nominal, data = full_summary)

tidy(model_gdp) |>
  gt() |>
  tab_header(
    title    = "Linear Regression Results",
    subtitle = "Response: Total Plastic Waste | Predictor: GDP Per Capita (Nominal)"
  ) |>
  tab_style(
    style     = cell_text(weight = "bold"),
    locations = list(cells_column_labels(),
                     cells_title())
  ) |>
  cols_label(
    term      = "Term",
    estimate  = "Estimate",
    std.error = "Std. Error",
    statistic = "t-stat",
    p.value   = "p-value"
  ) |>
  fmt_number(
    columns  = c(estimate, std.error, statistic),
    decimals = 4
  ) |>
  fmt_number(
    columns  = p.value,
    decimals = 4
  )
```

**Analysis:** From the summary tables that were constructed, it would appear that countries with higher levels of plastic pollution tend to be of a lower income level in terms of GDP per capita. The total plastic waste for countries with low GDP's per capita are much higher than that of countries with middle and high GDP's per capita.

The dumbbell plot seems to support this assumption, as it compares each country’s rank in GDP per capita (blue) with its rank in total plastic waste (orange), with gray lines showing the gap between the two. If higher GDP per capita led to more plastic waste, the points would be close together, but many countries show large differences. For example, Nigeria has a much higher plastic waste rank than its GDP per capita rank, while the United States and shows the opposite pattern, with a higher GDP per capita ranking but a lower plastic waste ranking. Many of the countries in the top 10 in plastic waste seem to be on the lower end of the rankings when it comes to GDP per capita, which would suggest that there could be a negative relationship present here.

However, the scatter plot challenges this assumption of a negative relationship. From the plot, Nigeria and the Philippines are responsible for the highest total plastic waste despite having GDP's per capita near zero, but they are the only points in that area, suggesting them to be potential outliers. There does not seem to be a relationship present for wealthier countries either, as countries like the United States and Germany which have significantly higher GDP's per capita, produce relatively moderate amounts of plastic waste. Many of the countries seem to have comparable levels of plastic waste at differing levels of GDP per capita.

Finally, from the simple regression linear model, which examines the relationship between GDP per capita and total plastic waste across countries, it is apparent that there is no correlation present between these two variables. The estimated slope was -0.0747, barely indicating a very slight negative relationship and the p-value of 0.7095 is extremely high, which indicates that there is no significant evidence for an association between GDP per capita and total plastic waste.

## Section 5: Final Conclusions

**RQ #1:** From the visualizations of the Sankey diagram and the heat map, it would appear that two countries, Nigeria and the Philippines, are outliers in the distribution of plastic waste across the globe, as the top companies responsible for plastic waste around the world seem to "spread the wealth" in terms of pollution among different countries.

**RQ #2:** Initially, the summary tables and the dumbbell plot would suggest that countries with a lower GDP per capita tend to have more plastic waste, but a closer inspection of the scatter plot and the linear regression model conclude that there is no observable relationship between the GDP per capita of a country and its plastic waste.
